Improving Software Maintainability through automatic refactoring of Code Clones

### 2019-10-26 saner/paper.tex

Most clone detection tools can be configured using thresholds. These thresholds indicate the minimum number of lines, tokens and/or statements that must be spanned for duplicate fragments to be considered clones. Often, such thresholds are intuitively chosen~\cite{li2006cp, roy2009mutation} or based on a quartile distribution of empirical data~\cite{alves2010deriving}. Using the maintainability score we can find support for which thresholds should be chosen to increase the chance to find clones that improve maintainability when refactored. %TODO: Check whether this really fits here

@@

\section{Survey existing clone detection tools}\label{sec:survey}
%To detect refactorable clones according to the definitions introduced in Sec.~\ref{sec:research}, we survey existing clone detection tools for their suitability.
To investigate the impact of the definitions introduced in Sec.~\ref{sec:research}, we need a tool to detect refactorable clones according to these definitions. We conducted a short survey on (recent) clone detection tools to detect such clones that can be refactored. The results of our survey are displayed in table~\ref{table:dettools}. We chose a set of tools that are open source and can analyze a popular object-oriented programming language. Next, we formulate the following four criteria by which we analyze these tools:
\begin{enumerate}
    \item \textbf{Should find clones in any context.} Some tools only find clones in specific contexts, such as only method level clones. We want to perform an analysis of all clones in projects to get a complete overview.
\item \textbf{Finds clone classes in control projects.} We assembled a number of control projects to assess the validity of the surveyed clone detection tools.
\item \textbf{Can analyze resolved symbols.} When detecting the types proposed in Sec.~\ref{sec:rtypes}, it is important that we can analyze resolved symbols (for instance a type reference). The rationale for this is further explained in~\ref{chap:challenge}.
\item \textbf{Extensive detection configuration.} Detecting our clone definitions, as proposed in Sec.~\ref{sec:rtypes}, require to have some understanding about the meaning of tokens in the source code (whether a certain token is a type, variable, etc.). The tool should recognize such structures, in order for us to configure our clone type definitions in the tool.
\end{enumerate}

@@

Based on these relations, CR determines where to place the cloned fragment when extracted to a new method. These categories are mutually exclusive: a clone class is flagged as the first relation in the above list that it applies to. The code of clones that have a \textit{Common Class} relation can be refactored by placing the cloned code in this same class~\cite{fowler2018refactoring}. The code of clones with a \textit{Common Hierarchy} relation can be placed in the intersecting class in the hierarchy (the class all clone instances have in common as an ancestor)~\cite{fowler2018refactoring}. The code of clones with a \textit{Common Interface} relation can be placed in the intersecting interface~\cite{mohnen2002interfaces}, but in the process has to become part of the classes' public contract. The code of clones that are \textit{Unrelated} can be placed in a newly created place. The state-of-the-art uses a utility class~\cite{mazinanian2016jdeodorant}, whereas CR creates a new superclass or interface abstraction. The reason CR creates a superclass or interface rather than a utility class is that it makes the relation explicit and allows previously unrelated clones to become related. % TODO: Elaborate on the superclass abstraction and interface (this is my contribution) --> in refactoring section

@@

\subsubsection{Refactorability} \label{sec:refactorability}
Several factors may obstruct the possibility to extract code to a new method~\cite{tsantalis2015assessing}:
\begin{itemize}
    \item \textbf{Complex Control Flow}: This clone contains \texttt{break}, \texttt{continue} or \texttt{return} statements, obstructing the possibility of method extraction.
    \item \textbf{Spans Part Of A Block}: This clone spans a part of a block statement. An example of this is shown in Fig.~\ref{fig:refactorabilitypartialblock}.
    \item \textbf{Is Not A Partial Method}: If the clone does not fall in the ``Partial method'' category of Sec.~\ref{sec:location}, the ``Extract Method'' refactoring technique cannot be applied.
    \item \textbf{Returns Multiple Values}: If a clone modifies or declares multiple pieces of data that it should return.
    \item \textbf{Top-Level Non-Statement}: If one of the top-level AST nodes of the clone is not a statement. For instance, if a (part of) an anonymous class is cloned.
    \item \textbf{Can Be Extracted}: This clone is a fragment of code that can directly be extracted to a new method. Then, based on the relation between the clone instances, further refactoring techniques can be used to refactor the extracted methods (for instance \todo{be consistent: ``Extract Method" or ``extract method"} ``pull up method'' for clones in sibling classes).
\end{itemize}
Clones that do not fall in the \textit{Can Be Extracted} category may require additional transformations or other techniques to refactor. CR only refactors the clones that \textit{Can Be Extracted}.

\subsubsection{AST Transformation} % TODO: Big todo --> Find literature, introduce proper reasoning why I need to do this.
CR uses JavaParser~\cite{smith20e17javaparser}: an AST-parsing library that allows to modify the AST and write it back to source code. To refactor clones, CR creates a new method declaration and moves all statements from a clone instance in the clone class to the new method. This method is placed according to the relation between the clone instances (see Sec.~\ref{sec:relation}). CR analyzes the source code of the extracted method and populates it with the following properties:
\begin{itemize}
  \item \textbf{Parameters}: For each variable used that is not accessible from the scope of the extracted method.
  \item \textbf{A return value}: If the method modifies or declares local data that is needed outside of its scope, or if the cloned fragments already returned data.
  \item \textbf{Thrown exception}: If the method throws an uncaught exception that is not a \texttt{RuntimeException}.
\end{itemize}
CR then removes all cloned code and replaces it with a call to the newly created method. In case of a return value, CR either assigns the call result, declares it or returns it accordingly.

@@

We also test the correctness of the refactored code using the JFreeChart project~\cite{jfreechart}. JFreeChart has a high test coverage and working tests, which allows us to test the correctness of the program after running CR.

@@

% TODO: Explain about the coupling metric. Discussion: it is not too relevant to the results because most refactorings are SAME CLASS which do not impact on the coupling metric. (maybe more about architectural refactorings because COUPLING, BALANCE, COHESION?)

>>>>

Most clone detection tools can be configured using thresholds. These thresholds indicate the minimum number of lines, tokens and/or statements that must be spanned for duplicate fragments to be considered clones. Often, such thresholds are intuitively chosen~\cite{li2006cp, roy2009mutation} or based on a quartile distribution of empirical data~\cite{alves2010deriving}. Using the maintainability score we can find support for which thresholds should be chosen to increase the chance to find clones that improve maintainability when refactored. \todo{Check whether this really fits here}

@@

\section{Survey existing clone detection tools}\label{sec:survey}
%To detect refactorable clones according to the definitions introduced in Sec.~\ref{sec:research}, we survey existing clone detection tools for their suitability.
To investigate the impact of the definitions introduced in Sec.~\ref{sec:research}, we need a tool to detect refactorable clones according to these definitions. We conducted a short survey on (recent) clone detection tools to detect such clones that can be refactored. The results of our survey are displayed in table~\ref{table:dettools}. We chose a set of tools that are open source and can analyze a popular object-oriented programming language. Next, we formulate the following four criteria by which we analyze these tools:
\begin{enumerate}
    \item \textbf{Should find clones in any context.} Some tools only find clones in specific contexts, such as only method level clones. We want to perform an analysis of all clones in projects to get a complete overview.
\item \textbf{Finds clone classes in control projects.} We assembled a number of control projects to assess the validity of the surveyed clone detection tools.
\item \textbf{Can analyze resolved symbols.} When detecting the clones proposed in Sec.~\ref{sec:research}, it is important that we can analyze resolved symbols (for instance a type reference).
\item \textbf{Extensive detection configuration.} Detecting our clone definitions, as proposed in Sec.~\ref{sec:research}, require to have some understanding about the meaning of tokens in the source code (whether a certain token is a type, variable, etc.). The tool should recognize such structures, in order for us to configure our clone type definitions in the tool.
\end{enumerate}

@@

Based on these relations, CR determines where to place the cloned fragment when extracted to a new method. These categories are mutually exclusive: a clone class is flagged as the first relation in the above list that it applies to. The code of clones that have a \textit{Common Class} relation can be refactored by placing the cloned code in this same class~\cite{fowler2018refactoring}. The code of clones with a \textit{Common Hierarchy} relation can be placed in the intersecting class in the hierarchy (the class all clone instances have in common as an ancestor)~\cite{fowler2018refactoring}. The code of clones with a \textit{Common Interface} relation can be placed in the intersecting interface~\cite{mohnen2002interfaces}, but in the process has to become part of the classes' public contract. The code of clones that are \textit{Unrelated} can be placed in a newly created place. The state-of-the-art uses a utility class~\cite{mazinanian2016jdeodorant}, whereas CR creates a new superclass or interface abstraction. The reason CR creates a superclass or interface rather than a utility class is that it makes the relation explicit and allows previously unrelated clones to become related.

@@

\subsubsection{Refactorability} \label{sec:refactorability}
Several factors may obstruct the possibility to extract code to a new method~\cite{tsantalis2015assessing}:
\begin{itemize}
    \item \textbf{Complex Control Flow}: This clone contains \texttt{break}, \texttt{continue} or \texttt{return} statements, obstructing the possibility of method extraction.
    \item \textbf{Spans Part Of A Block}: This clone spans a part of a block statement. An example of this is shown in Fig.~\ref{fig:refactorabilitypartialblock}.
    \item \textbf{Is Not A Partial Method}: If the clone does not fall in the ``Partial method'' category of Sec.~\ref{sec:location}, the ``Extract Method'' refactoring technique cannot be applied.
    \item \textbf{Returns Multiple Values}: If a clone modifies or declares multiple pieces of data that it should return.
    \item \textbf{Top-Level Non-Statement}: If one of the top-level AST nodes of the clone is not a statement. For instance, if a (part of) an anonymous class is cloned.
    \item \textbf{Can Be Extracted}: This clone is a fragment of code that can directly be extracted to a new method. Then, based on the relation between the clone instances, further refactoring techniques can be used to refactor the extracted methods (for instance ``Pull Up Method'' for clones in sibling classes).
\end{itemize}
Clones that do not fall in the \textit{Can Be Extracted} category may require additional transformations or other techniques to refactor. CR only refactors the clones that \textit{Can Be Extracted}.

\subsubsection{AST Transformation}
\todo{Big todo --> Find literature, introduce proper reasoning why I need to do this.}
CR uses JavaParser~\cite{smith2017javaparser}: an AST-parsing library that allows to modify the AST and write it back to source code. To refactor clones, CR creates a new method declaration and moves all statements from a clone instance in the clone class to the new method. This method is placed according to the relation between the clone instances (see Sec.~\ref{sec:relation}). CR analyzes the source code of the extracted method and populates it with the following properties:
\begin{itemize}
  \item \textbf{Parameters}: For each variable used that is not accessible from the scope of the extracted method.
  \item \textbf{A return value}: If the method modifies or declares local data that is needed outside of its scope, or if the cloned fragments already returned data.
  \item \textbf{Thrown exception}: If the method throws an uncaught exception that is not a \texttt{RuntimeException}.
\end{itemize}
CR then removes all cloned code and replaces it with a call to the newly created method. In case of a return value, CR either assigns the call result, declares it or returns it accordingly.

@@

We also test the correctness of the refactored code using the JFreeChart project~\cite{gilbert2002jfreechart}. JFreeChart has a high test coverage and working tests, which allows us to test the correctness of the program after running CR.

@@

\todo{Explain about the coupling metric. Discussion: it is not too relevant to the results because most refactorings are SAME CLASS which do not impact on the coupling metric. (maybe more about architectural refactorings because COUPLING, BALANCE, COHESION?)}
